---
title: "Remove bat roost signal"
author: "Anna Boser"
date: "2022-10-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stars) # ʻstarsʻ requires 'sf' 0.9.8 was found, but >= 1.0.3
#library(raster)
library(sf)
library(terra)
library(exactextractr)
library(here)
library(splines)
library(ggplot2)
library(data.table)
```

## Remove the signal from the bat roosts


```{r}
# bat occurance data
bat_occ <- rast(here("data","aggregated_data", "2017_aggregate_2017.tif"))

# roost location
bat_roosts <- st_read(here("data", "ca_bat_roosts", "ca_colonies.shp")) %>% st_make_valid()
```

```{r}
# turn bat occ data into a dataframe
bat_occ_df <- as.data.frame(bat_occ, xy=TRUE)
names(bat_occ_df) <- c("x", "y", "bats")

# x locations of roosts
x_roosts <- bat_roosts$xcoord

# y locations of roosts
y_roosts <- bat_roosts$ycoord
```

```{r}
# plot out the raw bat data

# ggplot() + 
#   geom_raster(data = bat_occ_df, aes(x=x, y=y, fill=bats)) + 
#   geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord))
```


```{r}
# spline <- lm(formula = bats ~ bs(x, degree = 3, knots = x_roosts)*bs(y, degree = 3, knots = y_roosts) , data = bat_occ_df)
# bat_occ_df$spline <- spline$fitted.values
```

```{r}
# ggplot() + 
#   geom_raster(data = bat_occ_df, aes(x=x, y=y, fill=spline)) + 
#   geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord))
```

That did not work. Second trial: plot out the spatial decay for each of the 8 roosts

```{r}
# Make a function that only keeps points that are within r from a point

# spot_filter <- function(d, x_center, y_center, r){
#   # first do a rough approximation 
#   d <- filter(d, x > x_center-r, x < x_center+r, y > y_center-r, y < y_center+r)
#   # then go over and get rid of the corners
#   d$dist = sqrt((d$x - x_center)^2+(d$y - y_center)^2)
#   d <- d[d$dist<r,]
#   return(d)
# }
# 
# d <- spot_filter(d=bat_occ_df, x_center=x_roosts[1], y_center=y_roosts[1], r=0.05)
```

```{r}
# ggplot(d) + 
#   geom_point(aes(x=dist, y=bats))
```

```{r}
# fit some exponential

# fit <- lm(log(bats) ~ log(dist), data = d)
# ggplot(d) + 
#   geom_point(aes(x=dist, y=bats)) + 
#   geom_line(aes(x=dist, y=exp(fit$fitted.values)), color = "red")
```

```{r}
# fit a logistic

# fit <- glm(bats/max(bats) ~ dist, family = "binomial", data = d)
# ggplot(d) + 
#   geom_point(aes(x=dist, y=bats)) + 
#   geom_line(aes(x=dist, y=fit$fitted.values*max(bats)), color = "red")
```

I thing the logistic regression wins. Neither is perfect though... maybe there's a better function out there. Now I need to code it up so that
1: I can get the logistic regressions for all 8 roosts
2: I can combine all of the regressions on the 3d plane

To combine all of the regressions, I think that I'll 
1: separately calculate the predictions over the entire study area for each regression
2: add them together
3: subtract the result from the observed bats

Ok full step by step: 
1: add a column for distance to each roost to bat_occ_df
2: Fit a separate logistic regression for each roost based on the closest 0.05 degrees
3: generate predictions for each roost regression
4: add together the predictions for each roost regression
5: subtract the total roost regression from the bats
6: plot out the result

```{r}
# 1: add a column for distance to each roost to bat_occ_df
add_dist_col <- function(d, roost_num){
  x_center = bat_roosts$xcoord[roost_num]
  y_center = bat_roosts$ycoord[roost_num]
  d[paste0("dist", roost_num)] = sqrt((d$x - x_center)^2+(d$y - y_center)^2)
  return(d)
}

for (i in 1:nrow(bat_roosts)){
  bat_occ_df <- add_dist_col(bat_occ_df, i)
}
```

```{r}
# 2: Fit a separate logistic regression for each roost based on the closest 0.05 degrees
# and
# 3: generate predictions for each roost regression

# logistic_pred <- function(d, roost_num, r){
#   d['dist'] <- d[paste0("dist", roost_num)]
#   save <- d
#   d <- d[d['dist']<r,]
#   max_bats = max(d$bats)
#   fit <- glm(bats/max_bats ~ dist, family = "binomial", data = d)
#   
#   plot <- ggplot(d) + 
#     geom_point(aes(x=dist, y=bats)) + 
#     geom_line(aes(x=dist, y=fit$fitted.values*max_bats), color = "red") + 
#     ggtitle(paste("logistic", bat_roosts$xcoord[roost_num], bat_roosts$ycoord[roost_num]))
#   print(plot)
#   
#   save[paste0("logistic", roost_num)] <- predict(fit, save)*max_bats
#   return(save)
# }

exponential_pred <- function(d, roost_num, r){
  d['dist'] <- d[paste0("dist", roost_num)]
  save <- d
  d <- d[d['dist']<r,]
  fit <- lm(log(bats + 1) ~ log(dist), data = d)
  
  plot <- ggplot(d) + 
    geom_point(aes(x=dist, y=bats)) + 
    geom_line(aes(x=dist, y=exp(fit$fitted.values)-1), color = "red") + 
    ggtitle(paste("exponential_r", r, "_", bat_roosts$xcoord[roost_num], bat_roosts$ycoord[roost_num]))
  print(plot)
  
  save[paste0("exponential_r", r, "_", roost_num)] <- exp(predict(fit, save))-1
  return(save)
}
```

```{r}
# for (i in 1:nrow(bat_roosts)){
#   bat_occ_df <- logistic_pred(d=bat_occ_df, roost_num=i, r=0.05)
# }
```

```{r}
for (i in 1:nrow(bat_roosts)){
  bat_occ_df <- exponential_pred(d=bat_occ_df, roost_num=i, r=0.05)
}
```

```{r}
far_from_roosts <- filter(bat_occ_df, 
                          dist1 > 0.01, 
                          dist2 > 0.01, 
                          dist3 > 0.01,
                          dist4 > 0.01,
                          dist5 > 0.01,
                          dist6 > 0.01,
                          dist7 > 0.01,
                          dist8 > 0.01)
```


```{r}
# 4: add together the predictions for each roost regression
# and 
# 5: subtract the total roost regression from the bats

corrected_pred <- function(d, func="logistic"){
  d[func] <- 0
  for (i in 1:nrow(bat_roosts)){
    d[func] <- d[func] + d[paste(func, i, sep = "_")]
  }
  d[func] = d[,func] - mean(d[,func]) # remove baseline bat presence from the model. 
  d[paste0(func, "_pred")] <- d["bats"] - d[func]
  return(d)
}

# bat_occ_df <- corrected_pred(bat_occ_df, func="logistic")
far_from_roosts <- corrected_pred(far_from_roosts, func="exponential_r0.05")
```

```{r}
# 6: plot out the result for logistic

# ggplot() + 
#   geom_raster(data = bat_occ_df, aes(x=x, y=y, fill=bats)) + 
#   geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
#   ggtitle("original_bats")
# 
# ggplot() + 
#   geom_raster(data = bat_occ_df, aes(x=x, y=y, fill=logistic)) + 
#   geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
#   ggtitle("logistic_predictions")
# 
# ggplot() + 
#   geom_raster(data = bat_occ_df, aes(x=x, y=y, fill=logistic_pred)) + 
#   geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
#   ggtitle("logistic_corrected_bats")
```

```{r}
# plot out results for exponential

ggplot() + 
  geom_raster(data = far_from_roosts, aes(x=x, y=y, fill=bats)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("original_bats") + 
  scale_fill_distiller(palette="Spectral", direction = 1)

ggplot() + 
  geom_raster(data = far_from_roosts, aes(x=x, y=y, fill=exponential_r0.05)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("exponential_corrected_bats (0.01 degrees from roosts)") + 
  scale_fill_distiller(palette="Spectral", direction = 1)

ggplot() + 
  geom_raster(data = far_from_roosts, aes(x=x, y=y, fill=exponential_r0.05_pred)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("exponential_corrected_bats (0.01 degrees from roosts)") + 
  scale_fill_distiller(palette="Spectral", direction = 1)

```

There are two problems with the above plots. 
1) the exponential predictions are negative on the edges, so we need to bump those up to zero
2) there are some negative bat predictions near the roots, so we need to bump those up to zero

```{r}

ggplot() + 
  geom_raster(data = far_from_roosts, aes(x=x, y=y, fill=bats)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("original_bats") + 
  scale_fill_distiller(palette="Spectral", direction = 1)

# first bump negative exponential predictions up to zero
far_from_roosts$noneg <- ifelse(far_from_roosts$exponential_r0.05<0, 0, far_from_roosts$exponential_r0.05)

ggplot() + 
  geom_raster(data = far_from_roosts, aes(x=x, y=y, fill=noneg)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("exponential_corrected_bats (0.01 degrees from roosts)") + 
  scale_fill_distiller(palette="Spectral", direction = 1)

# get new bat predictions
far_from_roosts$noneg_pred <- far_from_roosts$bats - far_from_roosts$noneg

ggplot() + 
  geom_raster(data = far_from_roosts, aes(x=x, y=y, fill=noneg_pred)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("exponential_corrected_bats (0.01 degrees from roosts)") + 
  scale_fill_distiller(palette="Spectral", direction = 1)

# then bump negative bat predictions up to zero
far_from_roosts$noneg_pred_noneg <- ifelse(far_from_roosts$noneg_pred<0, 0, far_from_roosts$noneg_pred)

ggplot() + 
  geom_raster(data = far_from_roosts, aes(x=x, y=y, fill=noneg_pred_noneg)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("exponential_corrected_bats (0.01 degrees from roosts)") + 
  scale_fill_distiller(palette="Spectral", direction = 1)
```


```{r}
library(raster)
# turn the corrected stuff back into a raster
r <- rasterFromXYZ(far_from_roosts[, c('x', 'y', 'noneg_pred_noneg')], crs = crs(bat_occ))

# save
writeRaster(r, here("data","bats_roost_adjusted", "bats_exponential_corrected_2017.tif"), overwrite = TRUE)
```

```{r}
# plot(rast(here("data","bats_roost_adjusted", "bats_exponential_corrected_2017.tif")))
```

