---
title: "Remove bat roost signal"
author: "Anna Boser"
date: "2022-10-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stars) # ʻstarsʻ requires 'sf' 0.9.8 was found, but >= 1.0.3
#library(raster)
library(sf)
library(terra)
library(exactextractr)
library(here)
library(splines)
library(ggplot2)
```

## Remove the signal from the bat roosts


```{r}
# bat occurance data
bat_occ <- rast(here("data","aggregated_data", "2017_aggregate_2017.tif"))

# roost location
bat_roosts <- st_read(here("data", "ca_bat_roosts", "ca_colonies.shp")) %>% st_make_valid()
```

```{r}
# turn bat occ data into a dataframe
bat_occ_df <- as.data.frame(bat_occ, xy=TRUE)
names(bat_occ_df) <- c("x", "y", "bats")

# x locations of roosts
x_roosts <- bat_roosts$xcoord

# y locations of roosts
y_roosts <- bat_roosts$ycoord
```

```{r}
# plot out the raw bat data
ggplot() + 
  geom_raster(data = bat_occ_df, aes(x=x, y=y, fill=bats)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord))
```


```{r}
spline <- lm(formula = bats ~ bs(x, degree = 3, knots = x_roosts)*bs(y, degree = 3, knots = y_roosts) , data = bat_occ_df)
bat_occ_df$spline <- spline$fitted.values
```

```{r}
ggplot() + 
  geom_raster(data = bat_occ_df, aes(x=x, y=y, fill=spline)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord))
```

That did not work. Second trial: plot out the spatial decay for each of the 8 roosts

```{r}
# Make a function that only keeps points that are within r from a point

spot_filter <- function(d, x_center, y_center, r){
  # first do a rough approximation 
  d <- filter(d, x > x_center-r, x < x_center+r, y > y_center-r, y < y_center+r)
  # then go over and get rid of the corners
  d$dist = sqrt((d$x - x_center)^2+(d$y - y_center)^2)
  d <- d[d$dist<r,]
  return(d)
}

d <- spot_filter(d=bat_occ_df, x_center=x_roosts[1], y_center=y_roosts[1], r=0.05)
```

```{r}
ggplot(d) + 
  geom_point(aes(x=dist, y=bats))
```

```{r}
# fit some exponential
fit <- lm(log(bats) ~ log(dist), data = d)
ggplot(d) + 
  geom_point(aes(x=dist, y=bats)) + 
  geom_line(aes(x=dist, y=exp(fit$fitted.values)), color = "red")
```

```{r}
# fit some exponential
fit <- glm(bats/max(bats) ~ dist, family = "binomial", data = d)
ggplot(d) + 
  geom_point(aes(x=dist, y=bats)) + 
  geom_line(aes(x=dist, y=fit$fitted.values*max(bats)), color = "red")
```

I thing the logistic regression wins. Neither is perfect though... maybe there's a better function out there. Now I need to code it up so that
1: I can get the logistic regressions for all 8 roosts
2: I can combine all of the regressions on the 3d plane

To combine all of the regressions, I think that I'll 
1: separately calculate the predictions over the entire study area for each regression
2: add them together
3: subtract the result from the observed bats

Ok full step by step: 
1: add a column for distance to each roost to bat_occ_df
2: Fit a separate logistic regression for each roost based on the closest 0.05 degrees
3: generate predictions for each roost regression
4: add together the predictions for each roost regression
5: subtract the total roost regression from the bats
6: plot out the result

```{r}
# 1: add a column for distance to each roost to bat_occ_df
add_dist_col <- function(d, roost_num){
  x_center = bat_roosts$xcoord[roost_num]
  y_center = bat_roosts$ycoord[roost_num]
  d[paste0("dist", roost_num)] = sqrt((d$x - x_center)^2+(d$y - y_center)^2)
  return(d)
}

for (i in 1:nrow(bat_roosts)){
  bat_occ_df <- add_dist_col(bat_occ_df, i)
}
```

```{r}
# 2: Fit a separate logistic regression for each roost based on the closest 0.05 degrees
# and
# 3: generate predictions for each roost regression

logistic_pred <- function(d, roost_num, r){
  d['dist'] <- d[paste0("dist", roost_num)]
  save <- d
  d <- d[d['dist']<r,]
  max_bats = max(d$bats)
  fit <- glm(bats/max_bats ~ dist, family = "binomial", data = d)
  
  plot <- ggplot(d) + 
    geom_point(aes(x=dist, y=bats)) + 
    geom_line(aes(x=dist, y=fit$fitted.values*max_bats), color = "red") + 
    ggtitle(paste("logistic", bat_roosts$xcoord[roost_num], bat_roosts$ycoord[roost_num]))
  print(plot)
  
  save[paste0("logistic", roost_num)] <- predict(fit, save)*max_bats
  return(save)
}

exponential_pred <- function(d, roost_num, r){
  d['dist'] <- d[paste0("dist", roost_num)]
  save <- d
  d <- d[d['dist']<r,]
  fit <- lm(log(bats) ~ log(dist), data = d)
  
  plot <- ggplot(d) + 
    geom_point(aes(x=dist, y=bats)) + 
    geom_line(aes(x=dist, y=exp(fit$fitted.values)), color = "red") + 
    ggtitle(paste("exponential", bat_roosts$xcoord[roost_num], bat_roosts$ycoord[roost_num]))
  print(plot)
  
  save[paste0("exponential", roost_num)] <- exp(predict(fit, save))
  return(save)
}
```

```{r}
for (i in 1:nrow(bat_roosts)){
  bat_occ_df <- logistic_pred(d=bat_occ_df, roost_num=i, r=0.05)
}
```

```{r}
for (i in 1:nrow(bat_roosts)){
  bat_occ_df <- exponential_pred(d=bat_occ_df, roost_num=i, r=0.05)
}
```

```{r}
# 4: add together the predictions for each roost regression
# and 
# 5: subtract the total roost regression from the bats

corrected_pred <- function(d, func="logistic"){
  d[func] <- 0
  for (i in 1:nrow(bat_roosts)){
    d[func] <- d[func] + d[paste0(func, i)]
  }
  d[paste0(func, "_pred")] <- d["bats"] - d[func]
  return(d)
}

bat_occ_df <- corrected_pred(bat_occ_df, func="logistic")
bat_occ_df <- corrected_pred(bat_occ_df, func="exponential")
```

```{r}
# 6: plot out the result
ggplot() + 
  geom_raster(data = bat_occ_df, aes(x=x, y=y, fill=bats)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("original_bats")

ggplot() + 
  geom_raster(data = bat_occ_df, aes(x=x, y=y, fill=logistic)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("logistic_predictions")

ggplot() + 
  geom_raster(data = bat_occ_df, aes(x=x, y=y, fill=logistic_pred)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("logistic_corrected_bats")
```

```{r}
ggplot() + 
  geom_raster(data = bat_occ_df, aes(x=x, y=y, fill=bats)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("original_bats")

ggplot() + 
  geom_raster(data = bat_occ_df, aes(x=x, y=y, fill=exponential)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("exponential_predictions")

ggplot() + 
  geom_raster(data = bat_occ_df, aes(x=x, y=y, fill=exponential_pred)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("exponential_corrected_bats")

far_from_roosts <- filter(bat_occ_df, 
                          dist1 > 0.01, 
                          dist2 > 0.01, 
                          dist3 > 0.01,
                          dist4 > 0.01,
                          dist5 > 0.01,
                          dist6 > 0.01,
                          dist7 > 0.01,
                          dist8 > 0.01)
ggplot() + 
  geom_raster(data = far_from_roosts, aes(x=x, y=y, fill=exponential_pred)) + 
  geom_point(data = bat_roosts, aes(x=xcoord, y=ycoord)) + 
  ggtitle("exponential_corrected_bats (0.01 degrees from roosts)")
```


